Estimating the distribution of block sizes as a function of average transaction rates

### 2015-10-22 paper/paper.tex
Made some final changes before the review and also added the pdf file to the repository

There is always some chance that a given block will fill up since arrival times are stochastic. Specifically, for the $k$'th block this probability is
\[
P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right).
\]
This probability depends crucially on the process that $x_\tau$ follows. If we assume that $x_{\tau}$ is deterministic with a constant transaction rate of $\gamma$ MB/min, then 
\[
x_{\tau_k} - x_{\tau_{k-1}} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/\gamma 10)
\]
and the probability of a full block is available in closed form as
\[
P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right) = \int_1^\infty 1/10\gamma e^{-1/10\gamma x}dx = e^{-1/10\gamma}.
\]

@@

For the full 5000 blocks, the posterior mean of $\alpha$ is 1.34 and the lower bound of the 95\% credible interval is 1.29, so we have strong evidence that $\alpha$ is nonzero and that a constant transaction rate is not appropriate. This is not surprising since Bitcoin usage has gone up over time, but on a short enough time scale usage may be approximately constant. A block is around 10 minutes on average, so 5000 blocks takes a little over a month on average. It is possible that on a short enough time-scale, constant transaction rates are more appropriate. So we fit the gamma distribution to 1000 blocks at a time --- representing about a week --- by breaking the original 5000 blocks into five groups. For each of these groups the 95\% credible interval does not contain one --- the lowest lower bound is 1.17. Similarly for groups of 500 blocks, each representing about 3.5 days, the lowest lower bound is about 1.08. So we can safely conclude that the transaction rate is not constant. A likely reason for this is seasonality in transaction rates --- Bitcoin trades are probably more likely to happen during certain times of day than others, based on when relevant markets are open and when the bulk of Bitcoin users are awake. The {\it mean} transaction rate averaging over e.g. seasonality is probably approximately constant for time-scales this small, however. In any case, transaction rates are nonconstant even on short time-scales, which motivates modeling them as such.

@@

The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how transaction data is distributed requires more work. Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official time-stamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median of local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node with a maximum offset of 70 minutes. Because of the nature of the blocks' time-stamps, occasionally a block has a time-stamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid time-stamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the blockchain, so the upper bound is not observed, but the lower bound is at our disposal.

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block arrives. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time between block $k$ and block $k-1$ so that $\tau_{k}=\tau_{k-1} + \delta_k$, and now let $t_k$ denote the observed arrival time in the blockchain. Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k}} - x_{\tau_{k-1}}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing stochastic process. Most of the interesting modeling choices come from modeling this stochastic process. Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider nonnegative infinite activity L\'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$. The following discussion is brief -- see e.g. \citet{barndorff2012basics} for a fuller treatment. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continuous time and has independent and stationary increments with $x_0=0$. Brownian motion is the most common example. A nondecreasing example is the gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$. The gamma process is called an infinite activity process since for any time $\tau$ and increment $\delta>0$, $x_{\tau + \delta} - x_{\tau} > 0$.

@@

Since the observation times are observed with error, it is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form for any $tau$ and $\delta>0$. This simplifies fitting the model because it allows us to write down a relatively simple exact likelihood function, facilitating both maximum likelihood estimation and Markov chain Monte Carlo methods. The parameters of the Gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the transaction data increments. We will use the gamma process since it is widely available in statistical packages and in particular is available in Stan \citep{stan-software:2015}, unlike the main alternative, the inverse Gaussian.

@@

Combining these pieces we obtain the following model. For block $k=1,2,\dots,K$
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k-1} + \delta_k} - x_{\tau_{k-1}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\gamma\delta_k,\beta)\\
\delta_{k} = \tau_{k} - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$. To complete the model we assume $\sigma$, $\beta$, and $\gamma$ are independent in the prior with $\sigma \sim Cauchy(0, 2.5)$, $\beta \sim G(1, 1)$, and $\gamma \sim LN(0, 10)$. Table \ref{tab:modelfit} contains the results of fitting the model.

@@

 Table \ref{tab:probsims} contains estimates of the probability of a filled block under a variety of assumptions, using parameter estimates of the posterior distribution of the model using 1000 contiguous blocks ending with block 303151. Probabilities in the first row were computed exactly using the cdf of the exponential distribution. Probabilities in the second and third rows were computed by simulating 800,000 blocks from the model at the specified values for $\gamma$ and $\beta$, where $\beta^{(0.5)}$ denotes the posterior median for $\beta$. Probabilities in the fourth row were computed by simulating 100 blocks from the model for each of the 8000 draws of from the posterior of $\beta$ and with $\gamma$ fixed at the specified values. Probabilities in the fifth and final row were computed by simulating 100 blocks for each of the 8000 draws from the posterior of $\beta$ and the shifted posterior of $\gamma$, where a draw from the shifted posterior of $\gamma$ is a draw from the posterior plus the difference between the specified $\mathrm{E}[\gamma]$ and $\gamma$'s posterior mean. 
% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Tue Oct 13 00:53:54 2015
\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrrrr}
  \hline
$E[\gamma] =$ & 0.02 & 0.03 & 0.04 & 0.05 & 0.06 & 0.07 & 0.08 & 0.09 \\ 
  \hline
$\beta = \infty$ & 0.0067 & 0.0357 & 0.0821 & 0.1353 & 0.1889 & 0.2397 & 0.2865 & 0.3292 \\ 
  $\beta = 3\times\beta^{(0.5)}$ & 0.0082 & 0.0387 & 0.0853 & 0.1383 & 0.1915 & 0.2420 & 0.2874 & 0.3293 \\ 
  $\beta = \beta^{(0.5)}$ & 0.0117 & 0.0446 & 0.0917 & 0.1435 & 0.1958 & 0.2446 & 0.2881 & 0.3300 \\ 
  $\beta \sim p(\beta|x,t)$ & 0.0119 & 0.0452 & 0.0919 & 0.1437 & 0.1954 & 0.2438 & 0.2892 & 0.3304 \\ 
  $(\beta,\gamma) \sim p(\beta,\gamma|x,t)$ & 0.0118 & 0.0448 & 0.0916 & 0.1433 & 0.1955 & 0.2442 & 0.2896 & 0.3301 \\ 
   \hline
\end{tabular}
\caption{Probability of a full block under a variety of parameter values. In the first row the transaction rate is constant, in the second through fourth rows the transaction rate is stochastic with fixed mean $\gamma$, and in the final row the transaction rate is stochastic and unknown with mean $\mathrm{E}[\gamma]$. Under all five regimes the expected mean transaction rate is the same --- it is the spread of the distribution of transaction rates and mean transaction rates that changes across regimes.}
\label{tab:probsims}
\end{table}

Under a fixed mean transaction rate, the probability of a full block increases as $\beta$ decreases, especially for smaller mean transaction rates. For larger mean transaction rates there is little effect on the probability the next block is full from decreasing $\beta$. This is expected since increasing the spread of transaction rates while holding the mean constant increases the variance of block sizes. When block sizes are on average much smaller than 1 MB, increasing the variance of block sizes puts significantly more probability mass above 1 MB, but when the mean is near but still below 1 MB it has a relatively small effect. What is unexpected is that no matter what the expected mean transaction rate is, the probability of a block filling up is largely unaffected by taking into account parameter uncertainty in $\beta$ or in both $\beta$ and $\gamma$, which can be seen from the final three lines of Table \ref{tab:probsims}.

@@

%% These estimates seem rather high considering no block have filled up yet. Transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block is over 0.1. Even at a constant transaction rate of 0.02, the probability of a full block is 0.067. In 100 blocks, which is approximately 17 hours, the probability of at least one block filling up is $1-(1-0.067)^{100}\approx 1$. Since no block has filled up yet, clearly something is happening that prevents blocks from filling up. One possibility is that transaction rates are usually low and in no danger of causing a full block, but occasionally there are large spikes in the rate. Another possibility is strategic action by Bitcoin traders and miners to prevent blocks from filling up. Traders may be waiting to trade until a new block replaces a nearly full block, which is not much of a cost since the wait is only about 10 minutes on average. Similarly miners have an incentive to keep blocks small since that reduces the change of an orphan block --- a mined block which is not accepted by the blockchain --- so the miner of the current block may be rejecting some transactions in order to prevent the limit from being reached.

>>>>

There is always some chance that a given block will fill up since arrival times are stochastic. Specifically, for the $k$'th block this probability is
\[
P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right).
\]
This probability depends crucially on the process that $x_\tau$ follows. If we assume that $x_{\tau}$ is deterministic with a constant transaction rate of $\gamma$ MB/min, then 
\[
x_{\tau_k} - x_{\tau_{k-1}} = \gamma\delta_k \stackrel{ind}{\sim} Exp(1/10\gamma)
\]
and the probability of a full block is available in closed form as
\[
P\left(x_{\tau_k} - x_{\tau_{k-1}} > 1\right) = \int_1^\infty \frac{1}{10\gamma}e^{-x/10\gamma}dx = e^{-1/10\gamma}.
\]

@@

For the full 5000 blocks, the posterior mean of $\alpha$ is 1.34 and the lower bound of the 95\% credible interval is 1.29, so we have strong evidence that $\alpha$ is nonzero and that a constant transaction rate is not appropriate. This is not surprising since Bitcoin usage has gone up over time. A block is around 10 minutes on average, so 5000 blocks takes a little over a month on average. It is possible that on a short enough time-scale, transaction rates are approximately constant. So we fit the gamma distribution to 1000 blocks at a time --- representing about a week --- by breaking the original 5000 blocks into five groups. For each of these groups the 95\% credible interval does not contain one --- the lowest lower bound is 1.17. Similarly for groups of 500 blocks, each representing about 3.5 days, the lowest lower bound is about 1.08. So we can safely conclude that the transaction rate is not constant. A likely reason for this is seasonality in transaction rates --- Bitcoin trades are probably more likely to happen during certain times of day than others, based on when relevant markets are open and when the bulk of Bitcoin users are awake. The {\it mean} transaction rate averaging over e.g. seasonality is probably approximately constant for time-scales this small, however. In any case, transaction rates are nonconstant even on short time-scales, which motivates modeling them as such.

@@

The first part of the model is determined for us since we know $\delta_k\stackrel{iid}{\sim} Exp(1/10)$ by the design of Bitcoin, but determining the how transaction data is distributed requires more work. Crucial constraints on the models we can fit come from the data obtained from the block chain. Since Bitcoin is a distributed network without a home node, there is no official time-stamp for when each block arrives on the chain. Instead, the block miner's local time is recorded. Two constraints on the reported time that the block was mined are built into the Bitcoin protocol. First, any block submitted to the chain with a local time less than the median of local times of the last 11 blocks is rejected. Second, any block with a local time greater than network adjusted time plus two hours is rejected, where network adjusted time is defined as the local time plus the median difference between local time and the time at all nodes connected to the local node with a maximum offset of 70 minutes. Because of the nature of the blocks' time-stamps, occasionally a block has a time-stamp which is earlier than one or more previous blocks in the chain. Essentially, we have measurement error on the measurement times, and the lower and upper bounds for valid time-stamps define lower and upper bounds for the measurement error distribution. Unfortunately network adjusted time is not recorded in the block chain, so the upper bound is not observed, but the lower bound is at our disposal.

The second major constraint that the block chain imposes on us is that transaction data is only measured once a block arrives. So at random measurement times which we only observe with error, we observe the amount of transaction data since the previous measurement time --- though we observe this without error. These two constraints motivate a class of models for modeling Bitcoin transaction data. Once again, let $\tau_k$ denote the actual time of the arrival of the $k$'th block, $\delta_k$ the amount of time between block $k$ and block $k-1$ so that $\tau_{k}=\tau_{k-1} + \delta_k$, and now let $t_k$ denote the observed arrival time in the block chain. Further, let $x_\tau$ denote the amount data in the entire chain at time $\tau$ so that $x_{\tau_{k}} - x_{\tau_{k-1}}$ denotes the amount of data in block $k$. Now suppose the amount of transaction data in MBs is distributed according to some nondecreasing stochastic process. Most of the interesting modeling choices come from modeling this stochastic process. Conventional Brownian motion is inappropriate since the increments should be nonnegative while geometric Brownian motion is a poor choice since it cannot start from zero. Instead, we consider nonnegative infinite activity L\'{e}vy processes, often called subordinators, with closed form solutions for $x_{\tau}$. The following discussion is brief -- see e.g. \citet{barndorff2012basics} for a fuller treatment. The stochastic process $\{x_{\tau}:\tau>0\}$ is a L\'{e}vy process if it is continuous time and has independent and stationary increments with $x_0=0$. Brownian motion is the most common example. A nondecreasing example is the gamma process where $x_{\tau + \delta} - x_{\tau} \sim G(\alpha \delta, \beta)$ for any time $\tau$ and positive increment $\delta$. The gamma process is called an infinite activity process since for any time $\tau$ and increment $\delta>0$, $x_{\tau + \delta} - x_{\tau} > 0$.

@@

Since the observation times are observed with error, it is attractive that we can write down the distribution of $x_{\tau + \delta} - x_\tau$ in closed form for any $tau$ and $\delta>0$. This simplifies fitting the model because it allows us to write down a relatively simple exact likelihood function, facilitating both maximum likelihood estimation and Markov chain Monte Carlo methods. The parameters of the gamma or inverse Gaussian distributions can also be further modeled as stochastic processes in order to capture e.g. seasonal variation or stochastic volatility, though most reasonable models on these parameters will not allow for closed form distributions of the transaction data increments. We will use the gamma process since it is widely available in statistical packages and in particular is available in Stan \citep{stan-software:2015}, unlike the main alternative, the inverse Gaussian.

@@

Combining these pieces we obtain the following model. For block $k=1,2,\dots,K$
\begin{align*}
t_k|t_{1:(k-1)}, \tau_{1:K} &\sim LN(\log(\tau_k) - \sigma^2/2, \sigma)\mathbbm{1}(t_k > m_k)\\
x_{\tau_{k-1} + \delta_k} - x_{\tau_{k-1}}|\tau_{1:K} & \stackrel{ind}{\sim} G(\beta\gamma\delta_k,\beta)\\
\delta_{k} = \tau_{k} - \tau_{k-1} &\stackrel{iid}{\sim} Exp(1/10)
\end{align*}
where $x\sim LN(\mu,\sigma)$ means $\log(x) \sim N(\mu,\sigma)$. To complete the model we assume $\sigma$, $\beta$, and $\gamma$ are independent in the prior with $\sigma \sim half-Cauchy(0, 2.5)$, $\beta \sim G(1, 1)$, and $\gamma \sim LN(0, 10)$. Note that the half-Caucy distribution is the Cauchy truncated to the positive real line. Table \ref{tab:modelfit} contains the results of fitting the model.

@@

 Table \ref{tab:probsims} contains estimates of the probability of a filled block under a variety of assumptions, using parameter estimates of the posterior distribution of the model using 1000 contiguous blocks ending with block 303151. Probabilities in the first row were computed exactly using the cdf of the exponential distribution. Probabilities in the second and third rows were computed by simulating 800,000 blocks from the model at the specified values for $\gamma$ and $\beta$, where $\beta^{(0.5)}$ denotes the posterior median for $\beta$. Probabilities in the fourth row were computed by simulating 100 blocks from the model for each of the 8000 draws of from the posterior of $\beta$ and with $\gamma$ fixed at the specified values. Probabilities in the fifth and final row were computed by simulating 100 blocks for each of the 8000 draws from the posterior of $\beta$ and the shifted posterior of $\gamma$, where a draw from the shifted posterior of $\gamma$ is a draw from the posterior plus the difference between the specified $\mathrm{E}[\gamma]$ and $\gamma$'s posterior mean. 
% latex table generated in R 3.2.2 by xtable 1.7-4 package
% Tue Oct 13 00:53:54 2015
\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrrrr}
  \hline
$E[\gamma] = \hat{\gamma} =$ & 0.02 & 0.03 & 0.04 & 0.05 & 0.06 & 0.07 & 0.08 & 0.09 \\ 
  \hline
$\beta = \infty$ & 0.0067 & 0.0357 & 0.0821 & 0.1353 & 0.1889 & 0.2397 & 0.2865 & 0.3292 \\ 
  $\beta = 3\times\beta^{(0.5)}$ & 0.0082 & 0.0387 & 0.0853 & 0.1383 & 0.1915 & 0.2420 & 0.2874 & 0.3293 \\ 
  $\beta = \beta^{(0.5)}$ & 0.0117 & 0.0446 & 0.0917 & 0.1435 & 0.1958 & 0.2446 & 0.2881 & 0.3300 \\ 
  $\beta \sim p(\beta|x,t)$ & 0.0119 & 0.0452 & 0.0919 & 0.1437 & 0.1954 & 0.2438 & 0.2892 & 0.3304 \\ 
  $(\beta,\gamma) \sim p(\beta,\gamma|x,t)$ & 0.0118 & 0.0448 & 0.0916 & 0.1433 & 0.1955 & 0.2442 & 0.2896 & 0.3301 \\ 
   \hline
\end{tabular}
\caption{Probability of a full block under a variety of parameter values --- $\gamma$ is the mean transaction rate and $\beta$ is a dispersion parameter so that the variance of transaction rates is $\gamma/\beta$. In the first row the transaction rate is constant at $\hat{\gamma}$, in the second through fourth rows the transaction rate is stochastic with fixed mean $\hat{\gamma}$, and in the final row the transaction rate is stochastic and unknown with mean $\hat{\gamma}$.}
\label{tab:probsims}
\end{table}

Under a fixed mean transaction rate, the probability of a full block increases as $\beta$ decreases, especially for smaller mean transaction rates. For larger mean transaction rates there is little effect on the probability the next block is full from decreasing $\beta$. This is expected since increasing the spread of transaction rates while holding the mean constant increases the variance of block sizes. When block sizes are on average much smaller than 1 MB, increasing the variance of block sizes puts significantly more probability mass above 1 MB, but when the mean is near but still below 1 MB it has a relatively small effect. What is unexpected is that no matter what the expected mean transaction rate is, the probability of a block filling up is largely unaffected by taking into account parameter uncertainty in $\beta$ or in both $\beta$ and $\gamma$, which can be seen from the final three rows of Table \ref{tab:probsims}.

@@

%% These estimates seem rather high considering no block have filled up yet. Transaction rates have been fluctuating around 0.05 for several months and even under constant transaction rates, the probability of a full block is over 0.1. Even at a constant transaction rate of 0.02, the probability of a full block is 0.067. In 100 blocks, which is approximately 17 hours, the probability of at least one block filling up is $1-(1-0.067)^{100}\approx 1$. Since no block has filled up yet, clearly something is happening that prevents blocks from filling up. One possibility is that transaction rates are usually low and in no danger of causing a full block, but occasionally there are large spikes in the rate. Another possibility is strategic action by Bitcoin traders and miners to prevent blocks from filling up. Traders may be waiting to trade until a new block replaces a nearly full block, which is not much of a cost since the wait is only about 10 minutes on average. Similarly miners have an incentive to keep blocks small since that reduces the change of an orphan block --- a mined block which is not accepted by the block chain --- so the miner of the current block may be rejecting some transactions in order to prevent the limit from being reached.
